# Removing and Adding Variable Names {#addnames}

Upon reading data into a data frame object, you may encounter situations in which it makes sense to remove the variable names (and not the data associated with the variable names) or to add or replace variables names. In this chapter, you will learn simple techniques for removing the variable names (i.e., column names) from a data frame object and adding (or replacing) variable names in a data frame object.

#### Video Tutorial

Link to Video Tutorial: XXXX

#### Functions & Packages Introduced

Function           | Package
------------------ | ------------------
`names`            | base `R`
`colnames`         | base `R`
`c`                | base `R`
`head`             | base `R`

#### Initial Steps
  
If you haven't already, save the file called **"PersData.csv"** into a folder that you will subsequently set as your working directory. Your working directory will likely be different than the one shown below (i.e., `"H:/RWorkshop"`). As a reminder, you can access all of the data files referenced in this book by downloading them as a compressed (zipped) folder from the my GitHub site: https://github.com/davidcaughlin/R-Tutorial-Data-Files; once you've followed the link to GitHub, just click "Code" (or "Download") followed by "Download ZIP", which will download all of the data files referenced in this book. For the sake of parsimony, I recommend downloading all of the data files into the same folder on your computer, which will allow you to set that same folder as your working directory for each of the chapters in this book.
  
Next, using the `setwd` function, set your working directory to the folder in which you saved the data file for this chapter. Alternatively, you can manually set your working directory folder in your drop-down menus by going to *Session > Set Working Directory > Choose Directory...*. Be sure to create a new R script file (.R) or update an existing R script file so that you can save your script and annotations. If you need refreshers on how to set your working directory and how to create and save an R script, please refer to [Setting a Working Directory](#setwd) and [Creating & Saving an R Script](#createRscript).

```{r, eval=FALSE}
# Set your working directory
setwd("H:/RWorkshop")
```

Next, read in the .csv data file called **"PersData.csv"** using your choice of read function. In this example, I use the `read_csv` function from the `readr` package [@R-readr]. If you choose to use the `read_csv` function, be sure that you have installed and accessed the `readr` package using the `install.packages` and `library` functions. *Note: You don't need to install a package every time you wish to access it; in general, I would recommend updating a package installation once ever 1-3 months.* For refreshers on installing packages and reading data into R, please refer to [Packages](#packages) and [Reading Data into R](#read).

```{r, eval=FALSE}
# Install readr package if you haven't already
# [Note: You don't need to install a package every 
# time you wish to access it]
install.packages("readr")
```
```{r}
# Access readr package
library(readr)

# Read data and name data frame (tibble) object
personaldata <- read_csv("PersData.csv")

# View the names of the variables in the data frame (tibble) object
names(personaldata)
```
```{r}
# View data frame (tibble) object
personaldata
```

As you can see from the output generated in your console, the `personaldata` data frame object contains basic employee demographic information. The variable names include: `id`, `lastname`, `firstname`, `startdate`, and `gender`. *Technically, the `read_csv` function reads in what is called a "tibble" object (as opposed to a data frame object), but for our purposes a tibble will behave similarly to a data frame. For more information on tibbles, check out Wickham and Grolemund's [-@wickham2017] chapter on tibbles: http://r4ds.had.co.nz/tibbles.html.*

## Remove Variable Names from a Data Frame Object

In some instances, you may wish to remove the variable names from a data frame. For example, I sometimes write (i.e., export) a data frame object I've been cleaning in R so that I may use the data file with the statistical software program called Mplus [@mplus]. Because Mplus doesn't accept variable names within its data files, I may drop the variable names from the data frame object prior to writing to my working directory. 
  
To remove variable names, just apply the `names` function with the data frame  name as the argument, and then use either the `<-` operator with `NULL` to remove the variable names. 

```{r}
# Remove variable names
names(personaldata) <- NULL

# View just the first 6 rows of the data frame object in Console
head(personaldata)
```

As you can see, the variable names do not appear in the overwritten `personaldata` data frame object.

## Add Variable Names to a Data Frame Object

In other instances, you might find yourself with a dataset that lacks variable names (or has variable names that need to be replaced), which means that you will need to add those variable names to the data frame. 
  
Let's work with the `personaldata` data frame object from the previous section for practice. To add variable names, we can use the `colnames` function from base R, and enter the name of the data frame as the argument. Using the `<-` operator, we can specify the variable names using the `c` (combine) function that contains a vector of variable names in quotation marks (`" "`) as the arguments. *Remember to type a comma (`,`) between the function arguments, as commas are used to separate arguments from one another when there are more than one.* Please note that the it's important that the vector of variable names contains the same number of names as the data frame object has columns.

```{r}
# Add (or replace) variable names to data frame object
colnames(personaldata) <- c("id", "lastname", "firstname", "startdate", "gender")

# View just the first 6 rows of data in Console
head(personaldata)
```

Now the data frame object has variable names!

## Summary

In this chapter, we reviewed how to remove variable names from a data frame object and how to add variable names to a data frame object using the `names`, `colnames`, and `c` functions, which all come standard with your base R installation.