# Read Data: Chapter Supplement {.unnumbered #read_supplement}

In addition to the `read_csv` function from the `readr` package covered in [Reading Data into R](#read), we can read .csv files into R using the `read.csv` function from base R and the `Read` function from the `lessR` package. 

#### Functions & Packages Introduced {.unnumbered}

Function           | Package
------------------ | ------------------
`read.csv`         | base `R`
`Read`             | `lessR`

#### Initial Steps {.unnumbered #initsteps_read_supp}

If required, please refer to the [Initial Steps](#initsteps_read) section from the chapter for more information on these initial steps.

```{r, eval=FALSE}
# Set your working directory
setwd("H:/RWorkshop")
```

## `read.csv` Function from Base R {.unnumbered #readcsv_supp}

The `read.csv` file comes standard with base R, which means that you don't need to install a package to access the function. As the function name implies, this function is used when the source data file is in .csv format. Typically, the `read.csv` function requires only a single argument within the parentheses, which will be the *exact* name of the data file enclosed with quotation marks; the file should be located your working directory folder. Remember, R is a language where case and space sensitivity matters when it comes to names; meaning, if there are spaces in your file name, there needs to be spaces when the file name appears in your R script, and if some letters are upper case in your file name, there needs to be corresponding upper-case letters in your R script. Let's practice reading in a file called **"PersData.csv"** by entering the exact name of the file followed by the .csv extension, all within in quotation marks. Remember, the file called **"PersData.csv"** should already be saved in your working directory folder (see [Initial Steps](#initsteps_read_supp)).

```{r}
# Read data from working directory
read.csv("PersData.csv")
```

As you can see, the data that appear in your Console contains only a handful of rows and columns; nonetheless, this gives you an idea of how the read.csv function works.
  
Often, you will want to assign your data frame to an object that is stored in your Global Environment for subsequent use. By creating a data frame object, you can manipulate and/or analyze the data within the object using a variety of functions (and without changing the data in the source file). To create a data frame object, we simply (a) use the same `read.csv` function from above, (b) add either a `<-` or `=` operator to the left of the `read.csv` function, and (c) create a name of our choosing for the data frame object by entering that name to the left of the `<-` or `=` operator. You can name your data frame object whatever you would like as long as it doesn't include spaces, doesn't start with a numeral, and doesn't include special characters like `*` or `-` (to name a few). I recommend choosing a name that is relatively short but descriptive, and that is not the same as another R function or variable name that you plan to use. Below, I name the new data frame object `personaldata`.

```{r}
# Read in data and name data frame object
personaldata <- read.csv("PersData.csv")
```

## `Read` Function from `lessR` Package {.unnumbered #readlessr_supp}

Just like the `read.csv` and `read_csv` functions, the `Read` function from the `lessR` package [@R-lessR2020] can read in .csv files; however, it can also read in other file formats like .xls/x, .sas7bdat (SAS), and .sav (SPSS). 
  
To use the `Read` function, the `lessR` package needs to be installed and accessed using the `install.packages` and `library` functions, respectively.

```{r, eval=FALSE}
# Install lessR package
install.packages("lessR")
```
```{r}
# Access lessR package
library(lessR)
```

When reading in a .csv file using the `Read` function, type the *exact* name of your data file from your working directory as an argument (followed by .csv and surrounded by quotation marks). Further, either the `<-` or `=` operator can be used to name the data frame object. 

```{r}
# Read data and assign to data frame object
personaldata <- Read("PersData.csv")
```

Let's print just the first six rows of the `personaldata` data frame object to the Console to verify that everything worked as intended.

```{r}
# Print just the first 6 rows of the data frame object in Console
head(personaldata)
```

For more information on the `Read` function from the `lessR` package, check out David Gerbing's website: http://www.lessrstats.com/videos.html.